Analytic Hierarchy Process: how to force consistency of pairwise comparisons matrix

### 2023-10-23 manuscript/FCPCOM/manuscript.tex
updating
updating papers
forcing consistency

\documentclass[authoryear,manuscript,12pt]{elsarticle}
\usepackage{setspace}
\doublespacing
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{rotating}
\usepackage{multirow}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{enumitem}

@@

\begin{abstract} 
Saaty's Analytic Hierarchy Process is an important method for assigning weight to multiple criteria. The logic of this method is not complicated. First a pairwise comparisons matrix is generated for the multiple criteria, and the normalized eigenvector of this pairwise matrix is used as the weight of the criteria. Despite this method represents an important inovation, its logic is not complicated at all. First a pairwise comparisons matrix is generated for the multiple criteria, and the normalized eigenvector of this matrix is used as the weight of criteria. But, since pairwise matrixes are usually generated manually and based only on some employee knowhow, there is a huge complexity on generating a consistent pairwise matrix. Especially when many criteria are used. This paper presents two algorithms that can be used to adjust inconsistent matrices, forcing such matrices to have a better consistency rate. The first method is a constructive method that uses the data inserted in the matrix to build a new improved one. The second method iteratively identifies inconsistencies, making minor changes in order to improve the matrix consistency rate. 
\end{abstract}
%%Graphical abstract
%\begin{graphicalabstract}
%\end{graphicalabstract}

@@

>>>>

\documentclass[authoryear,manuscript,12pt]{elsarticle}
\usepackage{setspace}
\doublespacing
\usepackage{algorithm}
\usepackage{algpseudocode}
%\usepackage{algorithm2e}
\usepackage{rotating}
\usepackage{multirow}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{enumitem}

@@

\begin{abstract} 
Saaty's Analytic Hierarchy Process is an important method for assigning weight to multiple criteria. Despite this method represents an important inovation, its logic is not complicated at all. First a pairwise comparisons matrix is generated for the multiple criteria, and the normalized eigenvector of this matrix is used as the weight of criteria. But, since pairwise matrixes are usually generated manually and based only on some employee knowhow, there is a huge complexity on generating a consistent pairwise matrix. Especially when many criteria are used. This paper presents two algorithms that can be used to adjust inconsistent matrices, forcing such matrices to have a better consistency rate. The first method is a constructive method that uses the data inserted in the matrix to build a new improved one. The second method iteratively identifies inconsistencies, making minor changes in order to improve the matrix consistency rate. 
\end{abstract}
%%Graphical abstract
%\begin{graphicalabstract}
%\end{graphicalabstract}

@@

\section{Algorithms for forcing matrix consistency}
\begin{algorithm}
\caption{function $consistencyRate()$}\label{alg:CA}
\begin{algorithmic}
\Require $[a_{ij}]_{i,j=1}^{n}, RI[x]_{x=1}^{10}$
\State{$\lambda_{max} = mainEigenvalue([a_{ij}]_{i,j=1}^{n})$ \footnote{main eigenvalue can be found using "Eigen/Eigenvalues library.}}
\State{$CI = (\lambda_{max} - n) / (n-1)$}
\State{$CR \gets CI/RI[n]$}
\State{\Return {$CR$}}	
\end{algorithmic}
\end{algorithm}

\subsection{Constructive algorithm}

\begin{algorithm}
\caption{function $constructivelyForceConsistency([a_{ij}]_{i,j=1}^{n})$}\label{alg:CA}
\begin{algorithmic}
\Require $[a_{ij}]_{i,j=1}^{n}$
\Ensure $consistencyRate([a_{ij}]_{i,j=1}^{n}) \leq 0.1$
\State{$CR = consistencyRate([a_{ij}]_{i,j=1}^{n})$}
\If {$CR \leq 0.1 $}
	\State{\Return {$[a_{ij}]_{i,j=1}^{n}$}}
\Else
	\For{$ 3 \leq k \leq n $}
		\State{$([a_{ij}]_{i,j=1}^{k}, CR) \gets forceConsistency([a_{ij}]_{i,j=1}^{k})$}
		\If{$CR \leq 0.1$}
			\State{\textbf{break for}}
		\EndIf
	\EndFor
\EndIf
\State{\Return {$([a_{ij}]_{i,j=1}^{n}, CR)$}}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{function $forceConsistency([a_{ij}]_{i,j=1}^{k})$}\label{alg:CA}
\begin{algorithmic}
\Require $[a_{ij}]_{i,j=1}^{n} \quad | \quad consistencyRate([a_{ij}]_{i,j=1}^{n-1}) \leq 0.1 $
\Ensure $consistencyRate([a_{ij}]_{i,j=1}^{n}) \leq 0.1$
\State{$CR \gets consistencyRate([a_{ij}]_{i,j=1}^{n})$}
\For{$j \leq n-2$}
	\For{$j+1 \leq k \leq n-1 $}
			\State{$g \gets 0$}
			\State{$s \gets 0$}
		\If{$a_{nj} > a_{nk} \textrm{ and } a_{jk} \leq 1 $}
			\State{$g \gets j$}
			\State{$s \gets k$}
		\ElsIf{$a_{nj} < a_{nk} \textrm{ and } a_{jk} \geq 1 $}
			\State{$g \gets k$}
			\State{$s \gets j$}
		\EndIf
		\If{$g \neq 0$}
			\State{$([a_{ij}]_{i,j=1}^{n} , CR) \gets reduce([a_{ij}]_{i,j=1}^{n}, g, s)$}
			\If{$CR \geq 0.1$}
				\State{$([a_{ij}]_{i,j=1}^{n} , CR) \gets encrease([a_{ij}]_{i,j=1}^{n}, g, s)$}
			\EndIf
		\EndIf
		\If{$CR \leq 0.1$}
			\State{\textbf{break for}}
		\EndIf
	\EndFor
	\If{$CR \leq 0.1$}
		\State{\textbf{break for}}
	\EndIf
\EndFor
\State{\Return {$([a_{ij}]_{i,j=1}^{n}, CR)$}}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{function $reduce([a_{ij}]_{i,j=1}^{n}, g, s)$}\label{alg:CA}
\begin{algorithmic}
\Require $g,s < n; [a_{ij}]_{i,j=1}^{n}$
\While{$a_{ng} > a_{ns}$}
	\State{$a_{ng} \gets a_{ng}^{--}$}
	\State{$a_{gn} \gets a_{gn}^{++}$}
	\State{$aux_{CR} = consistencyRate([a_{ij}]_{i,j=1}^{n})$}
	\If{$aux_{CR} < CR$}
		\State{$CR = aux_{CR}$}
		\If{$CR \leq 0.1$}
			\State{\textbf{break}}
		\EndIf
	\Else
		\State{$a_{ng} \gets a_{ng}^{++}$}
		\State{$a_{gn} \gets a_{gn}^{--}$}
		\State{\textbf{break}}
	\EndIf
\EndWhile
\State{\Return {$([a_{ij}]_{i,j=1}^{n}, CR)$}}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{function $encrease([a_{ij}]_{i,j=1}^{n}, g, s)$}\label{alg:CA}
\begin{algorithmic}
\Require $g,s < n; [a_{ij}]_{i,j=1}^{n}$
\While{$a_{ng} > a_{ns}$}
	\State{$a_{ns} \gets a_{ns}^{++}$}
	\State{$a_{sn} \gets a_{sn}^{--}$}
	\State{$aux_{CR} = consistencyRate([a_{ij}]_{i,j=1}^{n})$}
	\If{$aux_{CR} < CR$}
		\State{$CR = aux_{CR}$}
		\If{$CR \leq 0.1$}
			\State{\textbf{break}}
		\EndIf
	\Else
		\State{$a_{ns} \gets a_{ns}^{--}$}
		\State{$a_{sn} \gets a_{sn}^{++}$}
		\State{\textbf{break}}
	\EndIf
\EndWhile
\State{\Return {$([b_{ij}]_{i,j=1}^{n}, CR)$}}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{function $a_{ij}^{++}$}\label{alg:CA}
\begin{algorithmic}
\Require $a_{ij} \quad | \quad a_{ij} \in \{1/9, 1/7, 1/5, 1/3, 1, 3, 5, 7, 9\}$
	\If{$a_{ij} \geq 1 \textrm{ and } a_{ij} \neq 9 $}
		\State {$a_{ij} \gets a_{ij} + 2$}
	\ElsIf{$a_{ij} < 1$}
		\State {$a_{ij} \gets 1/(1/a_{ij} - 2)$}
	\EndIf
\State{\Return {$a_{ij}$}}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{function $a_{ij}^{--}$}\label{alg:CA}
\begin{algorithmic}
\Require $a_{ij} \quad | \quad a_{ij} \in \{1/9, 1/7, 1/5, 1/3, 1, 3, 5, 7, 9\}$
	\If{$a_{ij} \geq 3  $}
		\State {$a_{ij} \gets b_{ij} - 2$}
	\ElsIf{$a_{ij} \leq 1 \textrm{ and } a_{ij} \neq 1/9$}
		\State {$a_{ij} \gets 1/(1/a_{ij} + 2)$}
	\EndIf
\State{\Return {$a_{ij}$}}
\end{algorithmic}
\end{algorithm}

\subsection{Iterative algorithm}

\begin{algorithm}
\caption{function iterativelyForcingConsistency()}\label{alg:CA}
\begin{algorithmic}
\Require $\textrm{M}[a_{ij}]_{i,j=1}^{n}$
\Ensure $consistencyRate([b_{ij}]_{i,j=1}^{n}) \leq 0.1$

\State{\Return {$\textrm{M}[b_{ij}]_{i,j=1}^{n}$}}
\end{algorithmic}
\end{algorithm}
